Lifespan Analysis of Lemurs in Wild and Captive Environments

### 2024-12-15 paper/paper.qmd
prose refinedand frontend mentioned in paper

---
title: "Lifespan Analysis of Lemurs in Wild and Captive Environments"
subtitle: "Evaluating Effective Strategies and Highlighting Areas for Conservation Focus"
author: 
  - Khushaal Nandwani
thanks: "Code and data are available at: [https://github.com/khushaal-nandwani/predicting-lemurs/tree/main](https://github.com/khushaal-nandwani/predicting-lemurs/tree/main)."
date: today
date-format: long
abstract: "This study investigates the lifespan of lemurs in wild and captive environments using data from the Duke Lemur Center, aiming to identify how factors such as sex, species, genus, and birth month influence longevity. Using Generalized Linear Models, the analysis highlights that lemurs in captivity live significantly longer than their wild counterparts, emphasizing the impact of conservation practices. The findings show species-specific variations in lifespan and the pronounced influence of sex differences in captivity. These insights help identify effective practices in captivity and suggest areas for improvement to optimize conservation strategies, contributing to the survival of this endangered mammalian group."
format: pdf
number-sections: true
toc: true
bibliography: ref.bib
# editor:
#   render-on-save: true
---
```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)  
library(ggplot2)
library(dplyr)
library(patchwork) # For arranging plots side by side
library(arrow)
library(kableExtra)
library(rstanarm)
library(knitr)

@@

Lemurs are the world’s most endangered group of mammals (@citedukelemurcenter). Understanding the factors that influence their lifespan is crucial for guiding effective conservation efforts. The Duke Lemur Center (DLC), established in 1966, has played a pivotal role in the study and preservation of lemurs, maintaining the largest population of these species outside their native region Madagascar. Different factors can affect the longevity of lemurs in wild and captive. Longevity is important because the longer lemurs live, the more likely they are to reproduce and hence, contribute to the survival of their species.

My study uses a dataset from the DLC to analyze if and how sex, species, genus and month a lemur is born in can affect its lifespan in the wild and captivity. This study can provide insights into which lemur species we are struggling to conserve in captivity and their performance in the wild. It will highlight areas where our conservation efforts have been effective.

@@

The findings reveal that lemurs in captivity exhibit significantly longer lifespans. Males tend to live longer than females in both captive and wild environments, however the difference is more pronounced in captivity. Species-specific variations in lifespan were observed, with some species thriving in captivity while others perform better in the wild. Birth month effects were inconsistent, suggesting complex interactions with environmental factors.

@@

### Data Acquisition
 As @3zehr points out, DLC staff collected data about the lemurs according to standard operating procedures and USDA, AZA, and IACUC guidelines. They recorded information about births, deaths, weights, enclosure moves, behaviors, and other significant events on a daily basis. Originally, this data was stored in handwritten and typed paper formats. Later, it was computerized.

@@

### Data Processing
The DLC used SAS software to build a database for the lemur data. Data from various sources was imported into SAS Enterprise Guide, including ARKS, MedARKS, ZIMS, and spreadsheets. They wrote programs in SAS to extract, match, and join data, calculate new variables, and format the output. They also used tools within SAS Enterprise Guide Projects for calculations and formatting. The DLC uses a unique ID to match data for individual animals, and the taxonomic name for species-related variables. The data was validated by identifying and locating missing data, standardizing codes and text, investigating outliers, and comparing known values to the database output.Data that could not be verified was excluded from the published dataset.

@@

@@

For months in @fig-month-born-effects, we see captive lemurs outlive wild lemurs most of the months except August, January, March, Novemeber and September. It is worth exploring the reasons behind these differences by Duke Lemur Center, as they can provide insights into the impact of seasonal factors on lemur lifespan.

@@

## Key Findings 
One key insight is that captivity significantly increases lemur lifespan. Controlled environments provide consistent access to food, medical care, and protection from predators, reducing environmental stresses that contribute to shorter lifespans in the wild. This finding underscores the importance of captivity as a conservation tool for endangered species like lemurs.

Species-specific differences in lifespan emerge as a critical factor. Most of the species thrive in captivity due to tailored conservation practices, while some including `ALB`, `COR`, `MAC`, and `MON` perform better in the wild. Understanding the reasons behind these variations can inform targeted conservation efforts to improve outcomes.

@@

## Drawbacks of the Study
While the study provides valuable insights, it has its limitations. First, the dataset is skewed towards captive lemurs, limiting the robustness of wild lifespan analysis.

Second, the exclusion of interactions between birth type and other predictors in the combined model reduces the depth of insights.

@@

6. Using advanced modeling techniques like Bayesian models or neural networks for more nuanced predictions.

@@

The Duke Lemur Center employed standard operating procedures to gather data, adhering to USDA, AZA, and IACUC guidelines. The original data was handwritten and later digitized into two primary databases: the Animal Record Keeping System (ARKS) and MedARKS, and subsequently the Zoological Information Management System (ZIMS). These steps ensured consistency and traceability but may have introduced biases, particularly during digitization. More on this can be found in @sec-data_collection.

@@

The methodologies align with best practices in ecological studies, as outlined by Zehr et al. (2014), who emphasize the importance of longitudinal tracking and database integration for species conservation. The use of generalized linear models (GLMs) aligns with approaches recommended in similar studies to account for categorical and continuous predictors.

@@

The study’s observational nature limits causal inference but provides valuable insights into factors influencing lemur longevity.
Careful data cleaning and validation steps mitigate some biases, but limitations persist, particularly regarding the underrepresentation of wild populations.
Future work could employ advanced sampling techniques or collaborative data pooling with other conservation organizations to improve generalizability.

>>>>

---
title: "Lifespan Analysis of Lemurs in Wild and Captive Environments"
subtitle: "Evaluating Effective Strategies and Highlighting Areas for Conservation Focus"
author: 
  - Khushaal Nandwani
thanks: "Code and data are available at: [https://github.com/khushaal-nandwani/predicting-lemurs/tree/main](https://github.com/khushaal-nandwani/predicting-lemurs/tree/main)."
date: today
date-format: long
abstract: "This study investigates the lifespan of lemurs in wild and captive environments using data from the Duke Lemur Center, aiming to identify how factors such as sex, species, genus, and birth month influence longevity. Using Generalized Linear Models, the analysis found that lemurs in captivity live significantly longer than their wild counterparts, emphasizing the impact of conservation practices. The study also identified species-specific variations in lifespan and a pronounced influence of sex in captivity. The findings from this study can help determine what we are doing right and where we are going wrong."
format: pdf
number-sections: true
toc: true
bibliography: ref.bib
# editor:
#   render-on-save: true
---
```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)  
library(ggplot2)
library(dplyr)
library(patchwork) # For arranging plots side by side
library(arrow)
library(kableExtra)
library(rstanarm)
library(knitr)

@@

Lemurs are the world’s most endangered group of mammals (@citedukelemurcenter). Understanding the factors that influence their lifespan is important for guiding effective conservation efforts. The Duke Lemur Center (DLC), established in 1966, has played a pivotal role in the study and preservation of lemurs, maintaining the largest population of these species outside their native region Madagascar. Different factors can affect the longevity of lemurs in wild and captive. Longevity is important because the longer lemurs live, the more likely they are to reproduce and hence, contribute to the survival of their species.

My study uses a dataset from the DLC to analyze if and how sex, species, genus and month a lemur is born in can affect its lifespan in the wild and captivity. This study can help us understand which lemur species we are struggling to conserve in captivity and their corresponding performance in the wild. It will highlight areas where our conservation efforts have been effective.

@@

The findings show that lemurs in captivity exhibit significantly longer lifespans. Males tend to live longer than females in both captive and wild environments, however the difference is more pronounced in captivity. Species-specific variations in lifespan were observed, with some species thriving in captivity while others perform better in the wild. Birth month effects were inconsistent, suggesting complex interactions with environmental factors.

@@

### Data Acquisition {#sec-data-acquisition}
 As @3zehr points out, DLC staff collected data about the lemurs according to standard operating procedures and USDA, AZA, and IACUC guidelines. They recorded information about births, deaths, weights, enclosure moves, behaviors, and other significant events on a daily basis. Originally, this data was stored in handwritten and typed paper formats. Later, it was computerized.

@@

### Data Processing {#sec-data-processing}
The DLC used SAS software to build a database for the lemur data. Data from various sources was imported into SAS Enterprise Guide, including ARKS, MedARKS, ZIMS, and spreadsheets. They wrote programs in SAS to extract, match, and join data, calculate new variables, and format the output. They also used tools within SAS Enterprise Guide Projects for calculations and formatting. The DLC uses a unique ID to match data for individual animals, and the taxonomic name for species-related variables. The data was validated by identifying and locating missing data, standardizing codes and text, investigating outliers, and comparing known values to the database output.Data that could not be verified was excluded from the published dataset.

@@

## Model API and Frontend
The model API code along with its instructions is available on the GitHub repository [https://github.com/khushaal-nandwani/predicting-lemurs-api](https://github.com/khushaal-nandwani/predicting-lemurs-api). Please read the Readme file in the repository for more information on how to call the API.

There is also a frontend developed for the api, where one can dynamically enter the `sex`, `genus`, `species` and `month_born` of the lemur and can get predictions for that lemur's lifespan in both wild and captive enviorments. The frontend is live at [http://3.99.183.165:5173/](http://3.99.183.165:5173/). The code for the frontend is available at [https://github.com/khushaal-nandwani/predicting-lemurs-frontend](https://github.com/khushaal-nandwani/predicting-lemurs-frontend).

The API is built using FastAPI and deployed on an AWS EC2 Instance. The frontend app is made in Vue and deployed on the same EC2 Instance.

@@

For months in @fig-month-born-effects, we see captive lemurs outlive wild lemurs most of the months except August, January, March, Novemeber and September. It is worth exploring the reasons behind these differences by Duke Lemur Center, as they can help us understand the impact of seasonal factors on lemur lifespan.

@@

## Key Findings 
We learn that captivity significantly increases lemur lifespan. Controlled environments provide consistent access to food, medical care, and protection from predators, reducing environmental stresses that contribute to shorter lifespans in the wild. This finding underscores the importance of captivity as a conservation tool for endangered species like lemurs.

Species-specific differences in lifespan emerge as a factor. Most of the species thrive in captivity due to tailored conservation practices, while some including `ALB`, `COR`, `MAC`, and `MON` perform better in the wild. Understanding the reasons behind these variations can inform targeted conservation efforts to improve outcomes.

@@

## Drawbacks of the Study
Despite some useful findings, the study has its limitations. First, the dataset is skewed towards captive lemurs, limiting the robustness of wild lifespan analysis.

Second, the exclusion of interactions between birth type and other predictors was a tradeoff worth reconsidering.

@@

6. Using modeling techniques like Bayesian models or neural networks to improve predictive accuracy.

@@

The Duke Lemur Center employed standard operating procedures to gather data, adhering to USDA, AZA, and IACUC guidelines. The original data was handwritten and later digitized into two primary databases: the Animal Record Keeping System (ARKS) and MedARKS, and subsequently the Zoological Information Management System (ZIMS). These steps ensured consistency and traceability but may have introduced biases, particularly during digitization. More on how the data was acquired can be found in @sec-data-acquisition.

@@

The methodologies align with best practices in ecological studies, as outlined by Zehr et al. (2014), who emphasize the importance of longitudinal tracking and database integration for species conservation. The use of generalized linear models (GLMs) aligns with approaches recommended in similar studies to account for categorical and continuous predictors. More on how the data was processed can be found in @sec-data-processing.

@@

The study’s observational nature limits causal inference but helps us understand factors influencing lemur longevity.
Careful data cleaning and validation steps mitigate some biases, but limitations persist, particularly regarding the underrepresentation of wild populations.
Future work could employ other sampling techniques or collaborative data pooling with other conservation organizations to improve generalizability.
